\input{preamble}

\title{Building a Trustworthy Hubbard\\Exact-Diagonalization Solver\\
\large A step-by-step implementation tutorial}
\author{The \code{hubbard-ed} project}
\date{\today}

\begin{document}

\frontmatter
\hypersetup{pageanchor=false}
\maketitle
\cleardoublepage
\hypersetup{pageanchor=true}
\setcounter{page}{1}

\chapter*{How to use this tutorial}
\addcontentsline{toc}{chapter}{How to use this tutorial}

This tutorial develops a small exact-diagonalization solver for the
one-dimensional spin-$1/2$ Hubbard model.  The order is deliberate: each
chapter adds one abstraction only after the previous layer has a numerical
check.  The result is useful both as an educational implementation and as a
reference against which more ambitious representations can be tested.

Every complete program printed in this document lives in the adjacent
\code{code/} directory and is included with \code{\textbackslash
lstinputlisting}.  The manuscript therefore displays the same source that is
executed during validation.  From the repository root, run

\begin{verbatim}
for script in tutorial/code/step*.py; do
    .venv/bin/python "$script"
done
\end{verbatim}

The tutorial assumes familiarity with elementary second quantization and
linear algebra, but not with a many-body software library.  NumPy and SciPy are
the only numerical dependencies.

\tableofcontents

\mainmatter
\input{chapters/01_model_and_basis}
\input{chapters/02_fermionic_operators}
\input{chapters/03_sparse_hamiltonian}
\input{chapters/04_matrix_free_lanczos}
\input{chapters/05_observables}
\input{chapters/06_validation}
\input{chapters/07_general_hopping}
\input{chapters/08_research_extensions}

\appendix
\chapter{Package map}

\begin{center}
\begin{tabular}{ll}
\toprule
Module & Responsibility \\
\midrule
\code{basis.py} & Fixed-sector enumeration and index maps \\
\code{operators.py} & Creation, annihilation, and hopping signs \\
\code{hamiltonian.py} & CSR, matrix-free, and generic one-body matrices \\
\code{lanczos.py} & Lowest eigenpairs and residual diagnostics \\
\code{observables.py} & Local, correlation, and Fourier observables \\
\code{analytic.py} & Two-site and noninteracting reference results \\
\bottomrule
\end{tabular}
\end{center}

The production source is intentionally compact.  Read each module after its
corresponding chapter, then inspect the matching tests to see how the stated
invariants are enforced.

\end{document}
