\chapter{Assembling the sparse Hamiltonian}
\label{chap:sparse}

\section{Separate diagonal and hopping actions}

For a basis state $s=(b_\uparrow,b_\downarrow)$, the interaction is diagonal:
\begin{equation}
\bra{s}H_U\ket{s}=U\,
\operatorname{popcount}(b_\uparrow\mathbin{\texttt{\&}}b_\downarrow).
\label{eq:interactiondiag}
\end{equation}
Every allowed hopping term produces a target state in the same
$(N_\uparrow,N_\downarrow)$ sector.  Its matrix element is $-t$ times the sign
returned by the operator layer.

Open chains contain the unique undirected bonds
\begin{equation}
(0,1),(1,2),\ldots,(L-2,L-1).
\end{equation}
Periodic chains add $(L-1,0)$ when $L>2$.  The $L=2$ periodic chain deliberately
contains the same single edge as the open chain; adding it twice would double
the hopping and invalidate the standard two-site analytic result.  For every
undirected edge, both hopping directions and both spin species are applied.

\section{Column-wise construction}

Sparse assembly follows the definition of a matrix action.  For every basis
column $j$ representing $\ket{s_j}$:
\begin{enumerate}
\item append the interaction value from Eq.~\eqref{eq:interactiondiag} at
      $(j,j)$ if it is nonzero;
\item apply each directed spin-preserving hop;
\item map the resulting state $s_i$ to a row index $i$;
\item append the signed matrix element at $(i,j)$.
\end{enumerate}
The triplets $(i,j,H_{ij})$ are first accumulated as a coordinate matrix and
then converted to compressed sparse row (CSR) form.  Duplicate entries are
summed and explicit zeros are removed.

\lstinputlisting[caption={Building and checking a CSR Hamiltonian.},label={lst:sparse}]{code/step03_sparse_hamiltonian.py}

\section{Hermiticity as an implementation invariant}

For real $t$ and $U$, the result must be real symmetric.  More generally,
\begin{equation}
H=H^\dagger.
\end{equation}
This test catches missing reverse hops, inconsistent signs, incorrectly
duplicated bonds, and many indexing errors.  Check both open and periodic
boundaries and more than one particle sector.

Sparse storage changes the scaling from $O(\basisdim^2)$ dense entries to
roughly $O(zL\basisdim)$ possible transitions, where $z$ is a small
coordination-dependent factor.  However, Python lists used during coordinate
assembly have nontrivial overhead.  The implementation therefore has a
separate conservative dimension guard for explicit CSR construction.

\paragraph{Checkpoint.}
For tiny systems, compare the CSR matrix with its conjugate transpose exactly
up to numerical roundoff, inspect the atomic-limit diagonal, and verify at
least one periodic-boundary matrix element whose fermionic sign is negative.
